---
output: github_document
execute:
  echo: false
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  echo = FALSE,
  message = FALSE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  fig.height = 4,
  fig.width = 7,
  fig.dpi = 600
)
library(rnsf)
grants_this_year <- load_nsf_grants(update_current_year=FALSE)
most_recent <- max(as.Date(grants_this_year$date, format="%m/%d/%Y"))
```

# rnsf

Unofficial package to interface with NSF API

It also has abstracts, dates, and more information for all grants up to `r most_recent`. 

* Webpage with package information: <https://bomeara.github.io/rnsf/>
* Github page: <https://github.com/bomeara/rnsf/>

# Installation

**New (Oct 2026)** You can just install this package the way you would any R package on github:

```
remotes::install_github("bomeara/rnsf")
```

(You'll need to first install `remotes` (by doing `install.packages("remotes")`) if you don't already have it). 

Thanks to the [piggyback](https://github.com/ropensci/piggyback) package, the huge cached files are now stored as releases rather than within the package, making install easy. 

# Usage

There are three main ways to get data:

* Use `rnsf::nsf_return()` with various arguments to pull in data directly from NSF's API for awards. Good if you want to find info by keyword, state, or other fields.
    * For example, if we're curious about grants relevant to Yellowstone National Park, we could call `rnsf:nsf_return(keyword="Yellowstone")`. It will call the API multiple times until it has downloaded info on all grants with that keyword and returned a data.frame object.
* Use `rnsf::load_nsf_grants()` to load cached data (available [here](https://github.com/bomeara/rnsf/releases/tag/most-recent)) for years you specify. By default, it loads cached data from the most recent year. Good if you want to do an analysis across thousands of grants.
    * If we wanted to look at grants from 2020 to the present, we can call `rnsf::load_nsf_grants(year_start=2020)` and it will download the cached grants.
	* If we wanted to add any grants since I last updated the cache, we can call `rnsf::load_nsf_grants(year_start=2020, update_current_year=TRUE)`. This will re-download all grants from the current year.
* Use `data(grfp)` after loading the package to load information on all the Graduate Research Fellowship Program awards and honorable mentions (these are not available from NSF's API, so I had to download each year's pair of spreadsheets).

Below are some examples of the package's utility; see <https://github.com/bomeara/rnsf/blob/master/README.Rmd> for the details of the code to make the plots.

## Bergograms

Scientist Jeremy Berg often graphs federal funding over time (see <https://jeremymberg.github.io/jeremyberg.github.io/>). The very useful website [Grant Witness](https://grantwitness.org) has adopted these graphs, including invaluable updates of national funding and breakdowns by NSF division (see, for example, [here](https://grantwitness.org/nsf/analyses/agency-pulse-nsf-grants)). With the `rnsf` package we can look at finer detail. For example, the US Census splits the 50 US states into four different regions: Northeast, Midwest, South, and West. We can look at funding over time by region instead of nationally:

```{r bergogram, warning=FALSE, echo=FALSE, fig.width=8, fig.height=3}
library(tidyverse)
recent_grants <- rnsf::load_nsf_grants(year_start=2019)
recent_grants <- dplyr::left_join(recent_grants, data.frame(perfStateCode=state.abb, Region=state.region), by='perfStateCode')
recent_grants$Region <- gsub("North Central", "Midwest", as.character(recent_grants$Region))
recent_grants$date <- as.Date(recent_grants$date, format="%m/%d/%Y")
recent_grants$year <- format(recent_grants$date, "%Y")
recent_grants$dayofyear <- timeDate::dayOfYear(timeDate::timeDate(recent_grants$date))
recent_grants$estimatedTotalAmt <- as.numeric(recent_grants$estimatedTotalAmt)
recent_grants$count <- 1

# make useful colors
years <- sort(unique(recent_grants$year))
n_years <- length(years)
if (n_years >= 2) {
  n_gray <- max(0, n_years - 2)
  gray_vals <- if (n_gray>0) gray.colors(n_gray, start=0.85, end=0.4) else character(0)
  cols <- c(gray_vals, "blue", "red")
} else {
  cols <- if (n_years==1) "blue" else character(0)
}
names(cols) <- years

organized_grants <- recent_grants |> arrange(year, dayofyear) |> filter(!is.na(Region)) |> group_by(year, Region) |> mutate(estimatedTotalAmt_by_year = cumsum(estimatedTotalAmt), total_grants = cumsum(count)) |> ungroup()

g <- ggplot(organized_grants, aes(x=dayofyear, y=total_grants, colour=year)) + geom_line() + theme_minimal() + xlab("Day") + ylab("Number of grants awarded") + scale_x_date(date_breaks = "4 months", date_labels =  "%b") + facet_wrap(~Region,  scales="free_y", nrow=1) + scale_colour_manual(values=cols)

print(g)

```

Note some important differences between the plots above and those from Grant Witness: they typically plot by financial year, not calendar year; they also filter out transfer grants (a grant moves between PIs or institutions) and the code above does not do that.

## Topic frequency over time

The ozone hole was discovered in [1985](https://www.usatoday.com/story/news/nation/2025/05/19/what-happened-to-the-hole-in-the-ozone-layer/83644470007/): pollution led to a depletion of the ozone layer, which shields people (and other organisms) from a great deal of UV radiation. The world came together and passed the Montreal Protocol to limit the pollution, chlorofluorocarbons, that was causing the damage. The hole is healing, but still requires research and monitoring. We can see how NSF grants mentioning "ozone hole" changed over time (though not all grants mentioning "ozone hole" study this issue; some could be using it as a comparison to some other issue, for example). We can include a regression before and after 1985 and show the 95% CI for the proportion in each year (truncated by the y-axis limits).

```{r ozone}
#| code-fold: true
library(rnsf)
library(ggplot2)
library(timeDate)
library(dplyr)
library(viridis)
library(binom)

grants <- rnsf::load_nsf_grants(year_start=1980, year_end=1995)


grants$ozone_hole <- grepl("ozone hole", grants$abstractText, ignore.case=TRUE)
grants$date <- as.Date(grants$date, format="%m/%d/%Y")
grants$year <- as.numeric(format(grants$date, "%Y"))

grants_by_year <- grants |> group_by(year) |> summarize(ozone = sum(ozone_hole), all = n())  |> ungroup()
grants_by_year$ozone_percent <- 100 * grants_by_year$ozone / grants_by_year$all
grants_by_year$ozone_lower_percent <- 100 * binom::binom.confint(x=grants_by_year$ozone, n=grants_by_year$all, method="exact")$lower
grants_by_year$ozone_upper_percent <- 100 * binom::binom.confint(x=grants_by_year$ozone, n=grants_by_year$all, method="exact")$upper
grants_by_year$ozone_upper_percent_truncated <- sapply(grants_by_year$ozone_upper_percent, min, 2*max(grants_by_year$ozone_percent))
g <- ggplot(grants_by_year, aes(x=year, y=ozone_percent, group=1))+ geom_errorbar(aes(ymin=ozone_lower_percent, ymax=ozone_upper_percent_truncated), colour="gray", alpha=0.4, width=0) + geom_point() + theme_minimal() + xlab("Year") + ylab("Percentage of grants mentioning 'ozone hole'") + geom_smooth(data=subset(grants_by_year, year<1985), se=FALSE, method="lm") + geom_smooth(data=subset(grants_by_year, year>=1985), se=FALSE, method="lm")  + ylim(c(0, 2*max(grants_by_year$ozone_percent)))
print(g)
```

## Keywords

Let's look just at awards made in any program with "bio" or "systematics" or "evolution" in the name (biology grants) and see how words have changed between those that mention collections-related work (voucher, collection, field work, fieldwork, specimen) or those that mention AI (AI, artificial intelligence, LLM, large language model) in the abstract. First as a proportion of all the grants of either kind:

```{r systematics_bio}
#| code-fold: true

library(tidyr)
library(dplyr)
library(ggplot2)
library(scales)

grants <- rnsf::load_nsf_grants(year_start=2016, update_current_year=FALSE)
bio <- grants[grepl("SYSTEMATICS|BIO|EVOLUTION", grants$program, ignore.case=TRUE),]
AI_words <- paste0(c(" AI ", "Artificial Intelligence", " LLM ", "Large.Language Model"), collapse="|")
collections_words <- paste0(c("voucher", "collection", "field work", "fieldwork", "specimen"), collapse="|")
bio$AI <- grepl(AI_words, bio$abstractText, ignore.case=TRUE)
bio$Collections <- grepl(collections_words, bio$abstractText, ignore.case=TRUE)
bio_relevant <- subset(bio, AI | Collections)
bio_relevant$Type <- "Both"
bio_relevant$Type[!bio_relevant$AI] <- "Collections"
bio_relevant$Type[!bio_relevant$Collections] <- "AI"
bio_relevant$date_formatted <- as.Date(bio_relevant$date, format="%m/%d/%Y")
bio_relevant <- bio_relevant[!is.na(bio_relevant$date_formatted),]
bio_relevant$year <- as.numeric(format(bio_relevant$date_formatted, "%Y"))
bio_aggregate <- bio_relevant |> group_by(year, Type) |> summarise(count=n(), funds=sum(as.numeric(fundsObligatedAmt)), .groups = "drop_last")
g <- ggplot(bio_aggregate, aes(x=year, y=funds, fill=Type))  + xlim(2016,NA) + geom_area(position = "fill") + xlab("Year of grant") + ylab("Proportion of funds") + theme_minimal() + ggtitle("Biology grants: division of funding between AI and collections grants") + scale_fill_manual(values=c("firebrick1", "darkorchid", "dodgerblue"))
print(g)
```

And then the actual amount of funding awarded:

```{r systematics_bio_line}
g <- ggplot(bio_aggregate, aes(x=year, y=funds, colour=Type))  + geom_line() + xlab("Year of grant") + ylab("Total funds") + theme_minimal() + ggtitle("Biology grants: division of funding between AI and collections grants") + scale_fill_manual(values=c("firebrick1", "darkorchid", "dodgerblue")) + scale_y_continuous(labels = unit_format(unit = "M", scale = 1e-6))
print(g)
```

## Table of award info

We can also look at a table with the number, not total money, of grants by state or territory by academic semester, for example (only including this year up to the last cache of the data, and just showing the top few states by grants).

```{r grantsbyyear}
#| code-fold: true

library(tidyr)
library(knitr)

grants <- rnsf::load_nsf_grants(year_start=2024, update_current_year=FALSE)
grants$awardeeStateCode <- toupper(grants$awardeeStateCode) # handle some early data that uses lowercase
grants$academic_semester <- rnsf::date_to_academic_semester(as.Date(grants$date, format="%m/%d/%Y"))
grants_aggregated <- grants |> filter(academic_semester %in% apply(expand.grid(c(2024:2026), c(" Fall", "  Spring")), 1, paste0, collapse="")) |> group_by(academic_semester, awardeeStateCode) |> summarise(total_awarded = n(), .groups = "drop_last") |> ungroup() |> tidyr::pivot_wider(names_from = academic_semester, values_from=total_awarded, values_fill=0) |> dplyr::arrange(desc(`2024 Fall`))
colnames(grants_aggregated) <- gsub("  ", " ", colnames(grants_aggregated))
colnames(grants_aggregated)[1] <- "Area"
grants_aggregated[,1] <- rnsf::abbreviation_to_state(unname(unlist(grants_aggregated[,1])))
knitr::kable(head(grants_aggregated), row.names=FALSE)
```

## Comparisons by state

We can see how number of awards and total value of awards by state or territory versus the average so far this year compares to average funding at this point of the year for 2017-2024 (so it encompasses two different administrations). Note that the color scale is symmetric -- a reduction is orange, a gain is purple, and no change is white.

```{r recent}
#| code-fold: true

library(lubridate)
library(ggrepel)
library(tidyverse)
library(PresElectionResults)
library(sf)

grants <- rnsf::load_nsf_grants(year_start=2017, update_current_year=FALSE)

current_day_of_year <- lubridate::yday(Sys.Date())
grants_modified <- grants
grants_modified$awardeeStateCode <- toupper(grants_modified$awardeeStateCode) # handle some early data that uses lowercase
grants_modified$date_formatted <- as.Date(grants_modified$date, format="%m/%d/%Y")
grants_modified <- grants_modified[!is.na(grants_modified$date_formatted),]
grants_modified$day_of_year <- lubridate::yday(grants_modified$date_formatted)
grants_modified$year <- as.numeric(format(grants_modified$date_formatted, "%Y"))
grants_modified$funds <- as.numeric(grants_modified$fundsObligatedAmt)
grants_modified$counts <- 1
grants_recent <- grants_modified |> group_by(awardeeStateCode) |> summarize(most_recent = max(date_formatted), days_since_grant = Sys.Date() - max(date_formatted))

grants_2026 <- subset(grants_modified, year==2026)
grants_2017_2024 <- subset(grants_modified, year>= 2017 & year <=2024) # so we have 8 years of data across two admins
grants_2017_2024 <- subset(grants_2017_2024, day_of_year < current_day_of_year) # so we compare similar
grants_aggregated_2026 <- grants_2026 |> group_by(awardeeStateCode) |> summarize(funds_per_year_2026 = sum(funds), counts_per_year_2026 = sum(counts)) |> arrange(awardeeStateCode) |> ungroup()
grants_aggregated_2017_2024 <- grants_2017_2024 |> group_by(awardeeStateCode) |> summarize(funds_per_year_8 = sum(funds, na.rm=TRUE)/8, counts_per_year_8 = sum(counts, na.rm=TRUE)/8) |> arrange(awardeeStateCode) |> ungroup()
grants_comparison <- full_join(full_join(grants_aggregated_2017_2024, grants_aggregated_2026), grants_recent)
grants_comparison$funds_per_year_2026[is.na(grants_comparison$funds_per_year_2026)] <- 0
grants_comparison <- subset(grants_comparison, !is.na(grants_comparison$funds_per_year_8)) # rarely get funding
grants_comparison <- subset(grants_comparison, grants_comparison$funds_per_year_8>50000) # eliminate entities which have traditionally low, rare funding as they make it hard to see the wait times for the others 
grants_comparison$percentage_usual_funding_by_this_date <- 100*grants_comparison$funds_per_year_2026 / grants_comparison$funds_per_year_8
grants_comparison$percentage_usual_award_count_by_this_date <- 100*grants_comparison$counts_per_year_2026 / grants_comparison$counts_per_year_8


# devtools::install_github("jaytimm/PresElectionResults") # mostly for the state boundaries, though perhaps useful for other analyses
# Using example from https://github.com/jaytimm/American-political-data-and-R

grants_comparison$state_abbrev <- grants_comparison$awardeeStateCode 


mp <- PresElectionResults::xsf_TileOutv10 |>
  left_join(PresElectionResults::pres_by_state) |>
              filter(year == 2024) |> left_join(grants_comparison)  |> mutate(margins = republican - democrat)

min_award <- min(mp$percentage_usual_award_count_by_this_date)

mp$`% of normal award count` <- mp$percentage_usual_award_count_by_this_date


g <- ggplot(mp) +  
  geom_sf(aes(fill = `% of normal award count`),
           color = 'darkgray', lwd = .15) +
  geom_sf(data = PresElectionResults::xsf_TileInv10, 
          fill = NA, 
          show.legend = F, 
          color = NA, 
          lwd=.5) +
		

  geom_sf_text(data = PresElectionResults::xsf_TileInv10,
                          aes(label = state_abbrev),
                          size = 1.5,
                          color='black') +
  scale_fill_distiller(palette = "PuOr",  
                        limit = c(min_award, 200-min_award), direction=1) +
  theme_void() +   theme(legend.position="bottom") + ggtitle("Percent of usual NSF award count")
  
  

# grants_comparison$days_since_grant <- as.numeric(grants_comparison$days_since_grant)
# g <- ggplot(grants_comparison, aes(x=days_since_grant, y=percentage_usual_funding_by_this_date, label=awardeeStateCode)) + geom_text_repel(max.overlaps=50, alpha=0.8) + theme_minimal() + xlab("Days since most recent grant") + ylab("2026 percentage of usual funding by this date") + geom_point(alpha=0.5) + scale_x_continuous(trans='log1p', breaks=c(1, 7, 30, 90, 120, 365, 2*365), limits=c(0, NA)) 
print(g)
```

```{r statemoney}

min_award <- min(mp$percentage_usual_funding_by_this_date)
scale_limits <- c(min_award, 200-min_award)

mp$`% of normal funding` <- mp$percentage_usual_funding_by_this_date

big_jump <- mp[(mp$`% of normal funding`>max(scale_limits)),]

mp$`% of normal funding`[mp$`% of normal funding`>max(scale_limits)] <- max(scale_limits)


g <- ggplot(mp) +  
  geom_sf(aes(fill = `% of normal funding`),
           color = 'darkgray', lwd = .15) +
  geom_sf(data = PresElectionResults::xsf_TileInv10, 
          fill = NA, 
          show.legend = F, 
          color = NA, 
          lwd=.5) +
		

  geom_sf_text(data = PresElectionResults::xsf_TileInv10,
                          aes(label = state_abbrev),
                          size = 1.5,
                          color='black') +
  scale_fill_distiller(palette = "PuOr",  
                        limit = scale_limits, direction=1) +
  theme_void() +   theme(legend.position="bottom") + ggtitle("Percent of usual NSF funding awarded per year")
  
  

print(g)
```

`r ifelse(nrow(big_jump)>0, paste0("Note that ", paste(big_jump$state_abbrev, collapse=", "), " had an increase of ", paste(round(big_jump$percentage_usual_funding_by_this_date), collapse=", "), "% but was truncated at ", round(max(scale_limits)), " so that no change would remain at the center of the plot colors."), "")`

## Rolling window

How is NSF awarding grants over time? This uses a two week rolling interval, showing the average grants awarded per day in that interval. This can pick up things like temporary stoppages of awards.

```{r rolling}
#| code-fold: true

grants <- rnsf::load_nsf_grants(year_start=2020, update_current_year=FALSE)
grants_modified <- grants
grants_modified$date_formatted <- as.Date(grants_modified$date, format="%m/%d/%Y")
grants_modified <- grants_modified[!is.na(grants_modified$date_formatted),]
grants_modified$year <- as.numeric(format(grants_modified$date_formatted, "%Y"))
grants_modified <- grants_modified[order(grants_modified$date_formatted),]
grants_modified$funding <- as.numeric(grants_modified$fundsObligatedAmt)


min_date <- -14+as.Date("01/01/2020", format="%m/%d/%Y")
rolling_counts <- data.frame(end_date = as.Date(min_date:Sys.Date()), count=NA, amount=NA)
for (count_index in sequence(nrow(rolling_counts))) {
	focal_end_date <- rolling_counts$end_date[count_index]
	grants_focal <- subset(grants_modified, date_formatted>(focal_end_date-14) & date_formatted<=focal_end_date)
	rolling_counts$count[count_index] <- nrow(grants_focal)
	rolling_counts$amount[count_index] <- sum(grants_focal$funding, na.rm=TRUE)
}

rolling_counts$dayofyear <- timeDate::dayOfYear(timeDate::timeDate(rolling_counts$end_date))
rolling_counts$year <- format(rolling_counts$end_date, "%Y")

# create a color vector: shades of gray for all but the last two years, which are blue and red
years <- sort(unique(rolling_counts$year))
n_years <- length(years)
if (n_years >= 2) {
  n_gray <- max(0, n_years - 2)
  gray_vals <- if (n_gray>0) gray.colors(n_gray, start=0.85, end=0.4) else character(0)
  cols <- c(gray_vals, "blue", "red")
} else {
  cols <- if (n_years==1) "blue" else character(0)
}
names(cols) <- years

g <- ggplot(rolling_counts, aes(x=dayofyear, y=count/14, colour=year)) + geom_line() + theme_minimal() + xlab("End of 2 week period") + ylab("Grants awarded per day, two week rolling average") + scale_colour_manual(values=cols) + scale_y_continuous(trans='log1p', breaks=c(0, 1, 5, 10, 20, 40, 80, 160)) + scale_x_date(date_breaks = "1 months", date_labels =  "%b") + geom_point(data=rolling_counts[nrow(rolling_counts),]) 
print(g)
```

And rolling window not on a log scale:

```{r rolling_no_log}
#| code-fold: true

g <- ggplot(rolling_counts, aes(x=dayofyear, y=count/14, colour=year)) + geom_line() + theme_minimal() + xlab("End of 2 week period") + ylab("Grants awarded per day, two week rolling average") + scale_colour_manual(values=cols) + scale_x_date(date_breaks = "1 months", date_labels =  "%b") + geom_point(data=rolling_counts[nrow(rolling_counts),]) 
print(g)
```

## Wordclouds

This pulls in all grants mentioning the clade of ants (`ants <- rnsf::nsf_return(keyword="Formicidae"`) and then uses the `rnsf::nsf_wordcloud()` function to list the words by frequency.

```{r wordcloud}
ants <- rnsf::nsf_return(keyword="Formicidae")
nsf_wordcloud(ants$abstractText)
```

## GRFP data

The [NSF Graduate Research Fellowship Program](https://www.nsfgrfp.org) (GRFP) is one of NSF's oldest and most impactful programs: it provides funding for three years of study for people in grad school, giving them flexibility (no need for a research or teaching assistantship); the stipends are also often higher than those for most grad students. Every year, names, affiliations, and research areas of awardees (those receiving the money) and those with honorable mentions are released [here](https://www.research.gov/grfp/AwardeeList.do?method=loadAwardeeList) but you can only get one year at a time. I have manually downloaded them all and incorporated them into the package. To use:

```{r grfp}
library(rnsf)
data(grfp)
```

You can then plot information or do other analyses. For example, the number of awards in Mathematical Sciences over time:

```{r grfpplot}
#| code-fold: true
grfp_math <- grfp[grepl("Mathematical Sciences", grfp$Field.of.Study),] |> subset(award_level=="Awardee")
grfp_math$year <- as.numeric(grfp_math$year)
grfp_math_by_year <- grfp_math |> group_by(year) |> summarize(count=n()) |> ungroup()
g <- ggplot(grfp_math_by_year, aes(x=year, y=count, group=1)) + geom_line() + theme_minimal() + xlab("GRFP Year") + ylab("Number of awards in 'Mathematical Sciences'")
print(g)
```

And the frequency of different subfields of math, showing just the first twenty from the past ten years of awards:

```{r grfpsubfields}
#| code-fold: true
library(stringr)
subfields <- stringr::str_to_title(gsub("Mathematical Sciences - ", "", grfp_math$Field.of.Study))
popularity <- t(t(sort(table(subfields), decreasing=TRUE)))
knitr::kable(head(popularity, 20))
```

# Updating the package

First, update the package version in the DESCRIPTION.

Then the directory containing the package source:

```
library(rnsf)
grants_this_year <- nsf_update_cached_this_year() # perhaps worth doing a new nsf_get_all() after the beginning of the year
devtools::build_readme()
pkgdown::build_site()
system("git add */*")
system("git commit -m'automatic update of page and data' -a --no-verify") # ignore the readme warning
system("git push")
```

If new GRFP results are out, download them, retitle them as Awardee or HonorableMention with year and .tsv (i.e., "Awardee2030.tsv") and put them in the inst/extdata directory in the package source. Reinstall the package. Then,

```
library(rnsf)
grfp <- compile_grfp()
usethis::use_data(grfp, overwrite=TRUE)
```

Then do the usual git things. 
