---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# rnsf

Unofficial package to interface with NSF API

It also has abstracts, dates, and more information for all grants up to March 28, 2026. 

* Webpage with package information: <https://bomeara.github.io/rnsf/>
* Github page: <https://github.com/bomeara/rnsf/>

# Installation

```
install.packages("remotes")
remotes::install_github("bomeara/rnsf")
```

# Examples

## Getting cached data

```
library(rnsf)
data(grants)
```

This will give you a data.frame with all data: `head(grants)`

It has multiple date fields which are in "%m/%d/%Y" format (but everything is stored as a raw list, leaving to the user to process it). 

## New search

You can also do a new search. For example, to get all info on ants:

```{r ants, warning=FALSE}
library(rnsf)
ants <- rnsf::nsf_return(keyword="Formicidae")
```

And there is a function for doing a wordcloud:

```{r wordcloud, warning=FALSE}
library(rnsf)
nsf_wordcloud(ants$abstractText)
```

We can plot funding over time, inspired by Jeremy Berg's graphs on federal funding (see <https://jeremymberg.github.io/jeremyberg.github.io/>)

```{r bergogram, warning=FALSE}
library(rnsf)
library(ggplot2)
library(timeDate)
library(dplyr)
library(viridis)

data(grants) # or use grants <- rnsf::nsf_get_all() to get the most current list (will take hours) or grants <- nsf_update_cached() to update from the last cached version
grants$date <- as.Date(grants$date, format="%m/%d/%Y")
grants$year <- format(grants$date, "%Y")
grants$dayofyear <- timeDate::dayOfYear(timeDate::timeDate(grants$date))
grants$estimatedTotalAmt <- as.numeric(grants$estimatedTotalAmt)
grants <- grants |> arrange(year, dayofyear) |> group_by(year) |> mutate(estimatedTotalAmt_by_year = cumsum(estimatedTotalAmt)) |> ungroup()
grants <- grants[!is.na(grants$year),]
grants_recent <- subset(grants, year>=2020)
g <- ggplot(grants_recent, aes(x=dayofyear, y=estimatedTotalAmt_by_year/1e9, colour=year)) + geom_line() + theme_minimal() + xlab("Day of year") + ylab("Cumulative billions of dollars awarded") + scale_colour_viridis_d(option="turbo", begin=0.2)
print(g)
```

We can also look at a table with the number, not total money, of grants by state by year, for example (showing just the first few states):

```{r grantsbyyear}
library(tidyr)
library(knitr)
grants_aggregated <- grants |> filter(year > 2023) |> group_by(year, awardeeStateCode) |> summarise(total_awarded = n(), .groups = "drop_last") |> ungroup() |> tidyr::pivot_wider(names_from = year, values_from=total_awarded, values_fill=0) |> dplyr::arrange(desc(`2024`))
knitr::kable(head(grants_aggregated,10))
```